Modeling the near-wake of a vertical-axis cross-flow turbine with 2-D and 3-D RANS

### 2016-05-06 paper.tex
Text wrapping
Add edits from thesis chapter; resolves #38 resolves #37 resolves #34 resolves #36

Despite the development of many simple engineering models based on blade element
momentum or vortex methods, it remains difficult to predict the performance of
cross-flow turbines (CFTs, often vertical-axis turbines or VATs) in all
cases---namely when solidity, or blade chord-to-radius ratio is high---a common
characteristic of smaller rotors, and those designed for marine hydrokinetic
(MHK) applications. With computing power becoming evermore available and
affordable, computational fluid dynamics (CFD) has become an attractive method
for predicting the performance of cross-flow turbines since it only requires
that turbulence be modeled, i.e., the physics can be described from theoretical
true first principles by solving the Navier--Stokes equations using a boundary
layer resolving body-fitted grid. However, CFD can be computationally expensive
when done in three dimensions, which may be necessary in some cases.

@@

An overview of 3-D blade-resolved cross-flow turbine simulations reported in the
literature is presented in Table~\ref{tab:cfd-refs}. The $k$--$\omega$ SST RANS
turbulence model is shown to be a popular choice due to its success in
predicting flows with adverse pressure gradients and
separation~\cite{Menter2003}. Higher fidelity methods that resolve the large
scales of turbulence, such as large-eddy and detached-eddy simulation have also
been used. Note that for all the studies listed, model validation was either
done for performance or wake predictions---not both---and in some case omitted
entirely.

@@

To date, little computational work has been done to attempt to design arrays of
CFTs, despite their prospects for closer spacing compared with axial-flow
turbines (AFTs). For example, Araya et al.~\cite{Araya2014} modeled the flow
through a VAT array using ``leaky Rankine body'' potential flow singularities,
which was able to rank relative---though not absolut---performance of array
configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
simulate a farm of cross-flow turbines, though this was not validated with
experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a
group of cross-flow turbines, observing higher power output for a staggered
configuration, but also did not compare with experimental results. Giorgetti
et al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
validation was not performed. Li and Calisal \cite{Li2010} used a 3-D vortex
line method to show mutually improved power output from two adjacent turbines,
though the simulations over-predicted the effects compared with experiments.
Antheaume et al.~\cite{Antheaume2008} used a blade element approach
coupled with a 3-D RANS solver to also show how close spacing can improve power
output of CFTs.

In this study we set out to model the performance and near-wake of the high
solidity University of New Hampshire Reference Vertical Axis Turbine (UNH-RVAT)
using the Reynolds-averaged Navier--Stokes equations discretized via the finite
volume method. Though studies in the literature generally focus on predicting
the turbine loading and the local blade boundary layer, we seek to model both
this and the larger scale flow produced by the rotor, i.e., the near-wake, which
is of interest for this particular turbine since it has been shown
experimentally that the near-wake's momentum and energy transport processes are
dominated by vertical advection \cite{Bachant2015-JoT}. It logically follows
that a 2-D simulation, which omits the vertical dimension, would not correctly
predict wake recovery and turbine--turbine interaction. However, it is of
interest to determine how wrong a 2-D model may be, since the computational
feasibility is attractive.

It is assumed that if the turbine power output and near-wake predictions from
the numerical model match those of the experiments, the flow field can be
inspected in greater detail, i.e., that the experiments will have been
``interpolated.'' This will provide access to unmeasured quantities, e.g.,
pressure, and allow inspection into where the dominant flow structures
originate. This will ultimately help develop and evaluate low-order wake
generator models for use in turbine array modeling. In summary, the questions we
hope to answer here are:

@@

    \item Can 3-D RANS ``interpolate'' the experimental results and provide
    insight to develop new low-order wake generators to represent CFTs?

    \item Does 3-D RANS realize the correct proportions of wake recovery
    mechanisms, i.e., are the 3-D blade-resolved results a good ``target'' for
    those a low-order model should produce?
\end{enumerate}

@@

To close the RANS equations, two different turbulence models were
used---Menter's $k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras
(SA) one equation model \cite{Spalart1992}. Both closures use the eddy-viscosity
approach---the SA employing a single additional scalar transport equation and
the SST two. The SST model was chosen due to its prominence in the literature
for simulating separating flows, which we assumed to be present in the current
problem in the form of dynamic stall. The SA model was shown by Ferreira
et al.~\cite{Ferreira2007} to match experimental particle image
velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
low Reynolds number case ($5 \times 10^4$). Further justification for using the
SA model for this case comes from Crivellini and D'Alessandro
\cite{Crivellini2014}, where they successfully modeled the laminar separation
bubble and subsequent boundary layer transition to turbulence at Reynolds
numbers similar to those investigated here.

@@

Meshes were generated using \textit{OpenFOAM}'s \textit{blockMesh} and
\textit{snappyHexMesh} utilities. Mesh topology consists of a background
hexahedral mesh, which is refined in all three directions by a factor of 2 in a
rectangular region containing the turbine and near-wake (0.9 m upstream, 1.3 m
downstream, $\pm 0.9$ m cross-stream, and $\pm 0.8$ m vertically). Cells
adjacent to the turbine shaft and struts are refined by a factor of 4, while
cells adjacent to the blades are refined by a factor of 6. To capture the
boundary layer, 20 layers are added next to the blades with an expansion ratio
of 1.2. Overall mesh refinement is controlled by a single parameter---the number
of cells in the streamwise direction, $N_x$.

@@

Simulations were run using the \textit{pimpleDyMFoam} solver from the
open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
\textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
coupling and is compatible with dynamic meshes. An Euler scheme was used to
advance the simulation forward in time. The case files required to replicate the
simulations are available from \cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST,
    Bachant2016-UNH-RVAT-2D-OpenFOAM-SA, Bachant2016-UNH-RVAT-3D-OpenFOAM-SST,
    Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.

@@

Initial and boundary conditions were set to match those of the tow tank as well
as possible. The velocity at the inlet, bottom, and side walls was fixed to 1
m/s to match the tow tank case, while the top boundary condition was a slip
velocity condition. Pressure was held fixed at the outlet. Note that in 2-D the
top and bottom boundary conditions are ``empty,'' which is an OpenFOAM
convention to indicate two-dimensionality.

@@

Both the $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified
for convergence of the turbine mean power coefficient with respect to grid
spacing and time step using 2-D domains. The grid topology was fixed, but the
number of cells per domain length was scaled proportionally, maintaining the
same background mesh cell aspect ratio. Results for this parameter sweep are
shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
streamwise grid points $N_x = 70$ was chosen. This corresponds to a total cell
count of approximately $5 \times 10^4$ for the 2-D cases and 16 million for the
3-D case.

@@

Time step dependence was evaluated using the 2-D $N_x=70$ grid, the results from
which are shown in Figure~\ref{fig:2d-br-verification}. It was seen that the
Spalart--Allmaras model converged well with decreasing time step, leading to a
final time step of 0.001 s. The results from the SST model show a local minimum
at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
minimum was chosen as the final time step to run the simulations. Note that the
SST model's convergence behavior may be due to its specific implementation in
\textit{OpenFOAM}, and not indicative of the nature of the model equations.
Verification studies for CFTs with this level of detail in the literature are
not common, though the final time step is comparable to others
\cite{Balduzzi2016}.

@@

Turbine operation in all cases was simulated for 10 seconds, or approximately
six rotor revolutions. Computations for the 3-D cases were run on 192 processes
(24 nodes $\times$ 8 cores each) and took on the order of 1,000 CPU hours per
second of simulated time. The 2-D simulations were run on a single processor and
took on the order of one CPU hour per second.

Flow statistics were calculated over the time interval spanning $5$--$10$
seconds, or approximately three rotor revolutions, from time series downsampled
to 50 Hz. It is assumed that the downsampling frequency is high enough above the
blade passage frequency such that differences from the variance (for computing
turbulence statistics) in the original velocity will be negligible.

@@

Mean performance quantities---power and overall rotor drag coefficients---were
calculated for the interval after two full rotor revolutions, beyond which
values had reached an approximately periodic state. Predictions for both the
mean rotor power and drag coefficients are shown in
Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
significantly overpredict turbine loading and therefore mechanical power output,
which is likely due to their increased blockage ratio, unresolved blade end
effects, and lack of blade support struts.

The 3-D simulations fair better at predicting the experimental measurements. The
Spalart--Allmaras model's mean power coefficient was 0.24 while the SST's was
0.33, which represent a 6\% under- and 30\% overprediction, respectively,
compared with the experiments. The apparent overprediction of rotor drag
coefficient could be an effect of the experimental procedure, where the ``tare
drag'' from the turbine mounting structure was measured without a turbine
installed, then subtracted in post-processing. Technically the local flow field
will have changed, and hence the tare drag will change as well.

@@

Visualizations of the complex vorticity field generated by the turbine are
presented for the 2-D and 3-D Spalart--Allmaras cases in
Figure~\ref{fig:br-vorticity-3d} and Figure~\ref{fig:br-vorticity-3d},
respectively. It can be seen how the upstream blade---as it turns back into the
streamwise direction---is shedding a large amount of spanwise vorticity due to
the separated flow. In the 3-D case, strong tip vortices are also present, which
trace the ``contracting'' wake flow on the $-y$ side of the turbine associated
with the induced vertical velocity field. The 3-D dynamic stall vortex also
shows asymmetry about the $x$--$y$ mid-rotor plane; once again highlighting the
importance of three-dimensional effects on wake dynamics.

@@

Turbulence kinetic energy profiles are also shown in
Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy was calculated as
\begin{equation}
    k = k_{\mathrm{RA}} + \frac{1}{2} \left(
    \overline{U^\prime}^2 +
    \overline{V^\prime}^2 +
    \overline{W^\prime}^2 \right),
    \label{eq:k}
\end{equation}
where $U^\prime = U - \overline{U}$ and $k_{\mathrm{RA}}$ is the kinetic energy
calculated by the turbulence model, which is zero for the SA model.

Both Spalart--Allmaras cases do a poor job predicting the turbulence kinetic
energy in the flow, since it must be resolved as variance in the velocity field.
The 2-D SST model does a good job predicting the peak in $k$ at $y/R=-1$, though
is missing the smaller peak at $y/R=-1$. This is once again likely due to
blockage issues, where local tip speed ratio is decreased, increasing the
blades' instantaneous angle of attack at this location on the downstream
passage. In contrast, the 3-D SST model predicts the $+y$ peak in turbulence
kinetic energy very well, though the $-y$ peak magnitude is overpredicted by
about 30\%. We also see some smearing of $k$ across the center of the rotor,
which is likely due to exaggerated levels of the turbulent eddy viscosity.

@@

In order to visualize the mean velocity field, vector arrows for the mean
cross-stream and vertical components are superimosed on top of contours of the
streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
CFD models predict the general structure of the mean velocity well, though the
SA case has slightly larger vertical mean flows, which could be due to stronger
tip vortex generation, or lower diffusivity compared with the SST model.
Additional discrepancies between CFD and experiments may be due to the top slip
boundary condition versus the experiment's free surface.

@@

Turbulence kinetic energy contours for the experimental measurements and each
CFD case at $x/D=1$ are presented in Figure~\ref{fig:br-cfd-kcont}. As seen in
the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model is resolving very
little of the flow unsteadiness. In contrast, the SST model does a good job
predicting the locations and magnitudes of various peaks in $k$. These are
generated along the top of the turbine via tip vortex shedding, and the $-y$
side of the turbine via dynamic stall. We do however see the smearing effect
from the dynamic stall vortex centered around $z/H=0$, which is likely more of
an issue with wake evolution rather than wake generation.

@@

Turbulent transport and streamwise pressure gradient terms show the largest
discrepancy between results. The 3-D SST case, despite doing a good job
predicting turbulence kinetic energy, significantly overpredicts the turbulent
transport term, while other CFD cases are comparable with the experiment. This
seems to be balanced by a large adverse pressure gradient, which is also present
to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
cases create a wake where the pressure gradient is acting to slightly accelerate
the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
experiment, though the 3-D delayed detached eddy simulation (DDES) of Boudreau
and Dumas \cite{Boudreau2015} concur with the adverse pressure gradient
condition.

@@

The effectiveness at predicting the mean velocity field gives credibility to the
prospect of using the computed flow field to interpolate the experimental
results, such that the CFD results can be used as a target for a low-order wake
generator or force parameterization. These results may also help develop new tip
loss corrections for blade element type models, which currently only exist for
axial-flow rotors, since they provide access to the pressure and shear forces
over the entire blade surface, which were not measured experimentally.
Ultimately, however, the computational cost of 3-D simulations---about 10,000
CPU hours per operating case---may be too expensive to be used for CFT
engineering work, especially considering the uncertainty involved compared with
physical model studies. Since the 3-D wake results appeared to be only weakly
asymmetrical in the $x$--$y$ plane about $z/H=0$, it may be possible to reduce
computing power by only modeling the top half of the rotor and using a symmetry
boundary condition at the mid-plane. However, 3-D blade-resolved RANS will
likely not be practical for turbine array simulations for quite some time, and
until they are, low-order models will need to be developed and improved for this
purpose.

>>>>

Despite the development of many simple engineering models based on blade element
momentum or vortex methods, it remains difficult to predict the performance of
cross-flow turbines for all cases---namely when solidity, or blade
chord-to-radius ratio is high---a common characteristic of smaller rotors, and
those designed for marine hydrokinetic (MHK) applications. With computing power
becoming evermore available and affordable, CFD based on the Reynolds-averaged
Navier--Stokes (RANS) equations has become an attractive method for predicting
the performance of cross-flow turbines. Since a direct numerical simulation
(DNS) of all scales at realistic Reynolds numbers is not currently feasible,
turbulence be modeled. However, with an appropriate turbulence model,
blade-resolved RANS presents a more physically realistic first principles based
approach versus simpler models based on momentum theory or potential flow.
However, CFD can be computationally expensive when done in three dimensions,
which may be necessary in some cases.

@@

An overview of 3-D blade-resolved cross-flow turbine simulations reported in the
literature is presented in Table~\ref{tab:cfd-refs}. The $k$--$\omega$ SST RANS
turbulence model is shown to be a popular choice due to its success in
predicting flows with adverse pressure gradients and
separation~\cite{Menter2003}. Higher fidelity methods that resolve the large
scales of turbulence, such as large eddy and detached eddy simulation have also
been used. Note that for all the studies listed, model validation was performed
for either performance or wake predictions---not both---and in some case omitted
entirely.

@@

To date, little computational work has been done to attempt to design arrays of
CFTs, despite their prospects for closer spacing compared with axial-flow
turbines (AFTs). For example, Araya et al.~\cite{Araya2014} modeled the flow
through a VAT array using ``leaky Rankine body'' potential flow singularities,
which was able to rank relative---though not absolute---performance of array
configurations. Goude and Agren~\cite{Goude2010} used a 2-D vortex method to
simulate a farm of cross-flow turbines, though this was not validated with
experiments. Durrani et al.~\cite{Durrani2011} used 2-D CFD to model a group of
cross-flow turbines, observing higher power output for a staggered
configuration, but also did not compare with experimental results. Giorgetti et
al.~\cite{Giorgetti2015} took a similar approach for 2-D array analysis using
turbine pairs inspired by Dabiri~\cite{Dabiri2011}, but again experimental
validation was not performed. Li and Calisal \cite{Li2010} used a 3-D vortex
line method to show mutually improved power output from two adjacent turbines,
though the simulations over-predicted the effects by approximately 5\% compared
with experiments. Antheaume et al.~\cite{Antheaume2008} used a blade element
approach coupled with a 3-D RANS solver to also show how close spacing can
improve power output of CFTs.

In this study we set out to model the performance and near-wake of the high
solidity University of New Hampshire Reference Vertical-Axis Turbine (UNH-RVAT)
using the Reynolds-averaged Navier--Stokes equations using the open-source
finite volume CFD package OpenFOAM, version 2.3.x. Though studies in the
literature generally focus on predicting the turbine loading and the local blade
boundary layer, we seek to model both this and the larger scale flow produced by
the rotor, i.e., the near-wake, which is of interest for this particular turbine
since it has been shown experimentally that the near-wake's momentum and energy
transport processes are dominated by vertical advection \cite{Bachant2015-JoT}.
It logically follows that a 2-D simulation, which omits the vertical dimension,
would not correctly predict wake recovery and turbine--turbine interaction.
However, it is of interest to determine how wrong a 2-D model may be, since the
lower computational cost of 2-D simulations is attractive.

We seek to validate 2-D and 3-D blade-resolved RANS models against the UNH-RVAT
mechanical power and near-wake measurements. If the blade loading and velocity
in the near-wake match well enough, the flow field can potentially be inspected
in greater detail, i.e., we may be able to make observations of quantities not
measured experimentally, e.g., pressure, and gain greater insight into where the
dominant flow structures originate. This will ultimately help develop and
evaluate low-order wake generator models for use in turbine array modeling. In
summary, the questions we hope to answer here are:

@@

    \item Do the flow fields predicted by 3-D RANS match the experimental wake
    measurements well enough to provide insight for developing new low-order
    wake generators to represent CFTs?

    \item Does 3-D RANS realize the correct proportions of wake recovery
    mechanisms, i.e., are the 3-D blade-resolved results a good reference case
    or ``target'' for those a low-order model should produce?
\end{enumerate}

@@

To close the RANS equations, two different turbulence models were used: Menter's
$k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras (SA) one equation
model \cite{Spalart1992}. Both closures use the eddy-viscosity approach---the SA
employing a single additional scalar transport equation for an
eddy-viscosity-like quantity $\tilde{\nu}$ and the SST solving equations for the
turbulence kinetic energy $k$ and specific dissipation $\omega$. The SST model
was chosen due to its prominence in the literature for simulating separating
flows, which we assumed to be present in the current problem in the form of
dynamic stall. The SA model was shown by Ferreira et al.~\cite{Ferreira2007} to
match experimental particle image velocimetry (PIV) results for a CFT in dynamic
stall, though this was a somewhat low Reynolds number case ($Re_c \approx 6
\times 10^4$). Further justification for using the SA model for this case comes
from Crivellini and D'Alessandro \cite{Crivellini2014}, where they successfully
modeled the laminar separation bubble and subsequent boundary layer transition
to turbulence at Reynolds numbers similar to those investigated here.

@@

Meshes were generated using OpenFOAM's blockMesh and snappyHexMesh utilities.
Mesh topology consists of a background hexahedral mesh, which is refined in all
three directions by a factor of 2 in a rectangular region containing the turbine
and near-wake (0.9 m upstream, 1.3 m downstream, $\pm 0.9$ m cross-stream, and
$\pm 0.8$ m vertically). Cells adjacent to the turbine shaft and struts are
refined by a factor of 4, while cells adjacent to the blades are refined by a
factor of 6. To capture the boundary layer, 20 layers were added next to the
blades with an expansion ratio of 1.2. Overall mesh refinement is controlled by
a single parameter---the number of cells in the streamwise direction, $N_x$.

@@

Simulations were run using OpenFOAM's pimpleDyMFoam solver, which uses a hybrid
PISO-SIMPLE algorithm for pressure-velocity coupling and is compatible with
dynamic meshes. An Euler scheme was used to advance the simulation forward in
time. The case files required to replicate the simulations are available from
\cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST, Bachant2016-UNH-RVAT-2D-OpenFOAM-SA,
Bachant2016-UNH-RVAT-3D-OpenFOAM-SST, Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.

@@

Initial and boundary conditions were set to match those of the tow tank as well
as possible. The velocity at the inlet, bottom, and side walls was fixed to 1
m/s to match the tow tank case, while the top boundary condition was a slip
velocity condition. Pressure was held fixed at the outlet while a zero-gradient
boundary condition was applied at the inlet (a typical
velocity-inlet/pressure-outlet case). Note that in 2-D the top and bottom
boundary conditions are ``empty,'' which is an OpenFOAM convention to indicate
two-dimensionality.

@@

Both the $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified
for convergence of the turbine mean power coefficient with respect to grid
spacing and time step using 2-D domains. The grid topology was fixed, but the
number of cells per domain length was scaled proportionally, maintaining the
same background mesh cell aspect ratio. Results for this parameter sweep are
shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
streamwise grid points $N_x = 70$ was chosen, which resulted in a nondimensional
wall distance $y^+ = u^* y / \nu \sim 1$, where $u^*$ is the friction velocity,
defined as $u^*=\sqrt{\tau_w / \rho}$, where $\tau_w$ is the wall shear stress.
This mesh resolution gave a total cell count of approximately $5 \times 10^4$
for the 2-D cases and 16 million for the 3-D case. Sensitivity to domain length
was assessed as well, showing a 2\% increase in mean power coefficient for the
2-D SST case with the domain extended an additional two rotor diameters
downstream.

@@

Time step dependence was evaluated using the 2-D $N_x=70$ grid, the results from
which are shown in Figure~\ref{fig:2d-br-verification}. It was seen that the
Spalart--Allmaras model converged well with decreasing time step, leading to a
final time step of 0.001 s. The results from the SST model show a local minimum
at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
minimum was chosen as the final time step to run the simulations. Note that the
SST model's convergence behavior may be due to its specific implementation in
OpenFOAM, and not indicative of the nature of the model equations. Verification
studies for CFTs with this level of detail in the literature are not common,
though the final time step is comparable to others \cite{Balduzzi2016}.

@@

Turbine operation in all cases was simulated for 10 seconds, or approximately
six rotor revolutions. Computations for the 3-D cases were run on 192 processes
(24 nodes $\times$ 8 cores each on Sandia National Labs' Red Mesa high
performance computing cluster) and took on the order of 1,000 CPU hours per
second of simulated time. The 2-D simulations were run on a single processor and
took on the order of one CPU hour per second.

Performance statistics were computed after the first revolution onward and flow
statistics were calculated over the time interval spanning $5$--$10$ seconds, or
approximately three rotor revolutions, from time series downsampled to 50 Hz. It
is assumed that the downsampling frequency is high enough above the blade
passage frequency such that differences from the variance (for computing
turbulence statistics) in the original velocity will be negligible.

@@

Predictions for both the mean rotor power and drag coefficients are shown in
Figure~\ref{fig:br-cfd-perf-bar-chart}. In general, the 2-D CFD cases both
significantly overpredict turbine loading and therefore mechanical power output,
which is due to their increased blockage ratio, unresolved blade end effects,
and lack of blade support struts.

The 3-D simulations matched the experimental measurements more closely. The
Spalart--Allmaras model's mean power coefficient was 0.24 while the SST's was
0.33, which represent a 6\% under- and 30\% overprediction, respectively,
compared with the experiments. Some of the apparent overprediction of rotor drag
coefficient could also be an effect of the experimental procedure, where the
``tare drag'' from the turbine mounting structure was measured without a turbine
installed, then subtracted in post-processing. With the turbine installed, flow
past the mounting frame will be higher due to blockage, meaning the tare drag
would be underestimated, and the CFD results for $C_D$ should be slightly higher
than the experimental measurements.

@@

Visualizations of the complex vorticity field generated by the turbine are
presented for the 2-D and 3-D Spalart--Allmaras cases in
Figure~\ref{fig:br-vorticity-2d} and Figure~\ref{fig:br-vorticity-3d},
respectively. It can be seen how the upstream blade---as it turns back into the
streamwise direction---is shedding a large amount of spanwise vorticity due to
the separated flow. In the 3-D case, strong tip vortices are also present, which
trace the ``contracting'' wake flow on the $-y$ side of the turbine associated
with the induced vertical velocity field. The 3-D dynamic stall vortex also
shows asymmetry about the $x$--$y$ mid-rotor plane; once again highlighting the
importance of three-dimensional effects on wake dynamics.

@@

Turbulence kinetic energy profiles are also shown in
Figure~\ref{fig:br-cfd-profiles}. The turbulence kinetic energy for the unsteady
RANS models was calculated as
\begin{equation}
    k = k_{\mathrm{RA}} + \frac{1}{2} \left(
    \overline{U^\prime}^2 +
    \overline{V^\prime}^2 +
    \overline{W^\prime}^2 \right),
    \label{eq:k}
\end{equation}
where $U^\prime = U - \overline{U}$ (the resolved velocity fluctuations) and
$k_{\mathrm{RA}}$ is the kinetic energy calculated by the turbulence model,
which is zero for the SA model.

Both Spalart--Allmaras cases did a poor job predicting the turbulence kinetic
energy in the flow, since it must be resolved as variance in the velocity field.
The 2-D SST model did a good job predicting the peak in $k$ at $y/R=-1$, though
is missing the smaller peak at $y/R=-1$. This is once again likely due to
blockage issues, where local tip speed ratio is decreased, increasing the
blades' instantaneous angle of attack at this location on the downstream
passage. In contrast, the 3-D SST model predicts the $+y$ peak in turbulence
kinetic energy very well, though the $-y$ peak magnitude is overpredicted by
about 30\%. We also see some smearing of $k$ across the center of the rotor,
which is likely due to exaggerated levels of the turbulent eddy viscosity.

@@

In order to visualize the mean velocity field, vector arrows for the mean
cross-stream and vertical components are superimposed on top of contours of the
streamwise component at $X/D=1$ in Figure~\ref{fig:br-cfd-mean-velocity}. Both
CFD models predict the general structure of the mean velocity well, though the
SA case has a slightly larger vertical mean flow component, which could be due
to stronger tip vortex generation, or lower diffusivity compared with the SST
model.

@@

Turbulence kinetic energy contours for the experimental measurements and each
CFD case at $x/D=1$ are presented in Figure~\ref{fig:br-cfd-kcont}. As seen in
the profiles in Figure~\ref{fig:br-cfd-profiles}, the SA model was not able to
resolve the majority of the unsteadiness in the flow. In contrast, the SST model
did a good job predicting the locations and magnitudes of various peaks in $k$.
These are generated along the top of the turbine via tip vortex shedding, and
the $-y$ side of the turbine via dynamic stall. We do however see the smearing
effect from the dynamic stall vortex centered around $z/H=0$, which is likely
more of an issue with wake evolution rather than wake generation.

@@

Turbulent transport and streamwise pressure gradient terms show the largest
discrepancy between results. The 3-D SST case, despite doing a good job
predicting turbulence kinetic energy, significantly overpredicts the turbulent
transport term, while other CFD cases are comparable with the experiment. This
seems to be balanced by a large adverse pressure gradient, which is also present
to a smaller degree in the 3-D SA case. Interestingly, in contrast, both 2-D CFD
cases create a wake where the pressure gradient is acting to slightly accelerate
the flow at $x/D=1$. Unfortunately, pressure data were not acquired from the
experiment, though results from the 3-D delayed detached eddy simulation (DDES)
of Boudreau and Dumas \cite{Boudreau2015} concur with the adverse pressure
gradient condition.

@@

The effectiveness at predicting the mean velocity field gives credibility to the
prospect of using the computed flow field to extrapolate the experimental
results, such that the CFD results can be used as a target for a low-order wake
generator or force parameterization. These results may also help develop new tip
loss corrections for blade element type models, which currently only exist for
axial-flow rotors, since they provide access to the pressure and shear forces
over the entire blade surface, which were not measured experimentally.
Ultimately, however, the computational cost of 3-D simulations---about 1,000 CPU
hours per simulated second (or roughly 10,000 CPU hours per operating
point)---may be too expensive to be used for CFT engineering work, especially
considering the uncertainty involved compared with physical model studies. Since
the 3-D wake results appeared to be only weakly asymmetrical in the $x$--$y$
plane about $z/H=0$, it may be possible to reduce computing power by only
modeling the top half of the rotor and using a symmetry boundary condition at
the mid-plane. However, 3-D blade-resolved RANS will likely not be practical for
turbine array simulations for quite some time, and until they are, low-order
models will need to be developed and improved for this purpose.
